---
output: html_document
editor_options: 
  chunk_output_type: console
---
## Red Wine Quality Analysis by Travis Moore

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r wine, include=FALSE, warning=FALSE}
getwd()
setwd('C:/Users/moore/C751Wine')
library(ggplot2)
wine <- read.csv('wineQualityReds.csv')

```

##### This dataset contains nearly 1600 observations across 13 variables. The goal of this analysis is to understand what factors, if any, contribute to wine quality ratings.

```{r initExplore, warning=FALSE}

names(wine)

```

### Histogram of Wine Quality Rating
```{r ratingHist, warning=FALSE}

ggplot(aes(x= quality), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous() +
  xlab("Quality Rating (0-10)") + ylab("Count")

summary(wine$quality)

```

##### I chose a bar chart so we can easily what the wines in the dataset are rated. A majority of our wines' ratings were between 5 and 8.

### Alcohol Content
#### Bar Chart
```{r}
ggplot(aes(x= alcohol), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous() +
  xlab("Alcohol % by volume") + ylab("Count")

```

##### I chose a bar chart so we can easily what the alcohol content of wines in the dataset.

#### Box Plot with Category Summaries
```{r alcohol, warning=FALSE}

ggplot(aes(factor(quality), 
            alcohol), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Alcohol % by volume") + ylab("Count")

```

##### The higher the alcohol content the higher the rating.

### Fixed Acidity
#### Bar Chart
```{r}
ggplot(aes(x= fixed.acidity), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous()+
  xlab("Fixed Acidity (tartaric acid - g / dm^3)") + ylab("Count")

```

##### I chose a bar chart so we can easily what the fixed acidity content of wines in the dataset.

#### Box Plot with Category Summaries
```{r facid, warning=FALSE}

ggplot(aes(factor(quality), 
            fixed.acidity), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  xlab("Fixed Acidity (tartaric acid - g / dm^3)") + ylab("Count")

```

##### The higher the rating, the higher the fixed acidity of the wine.

### Citric Acid
#### Bar Chart
```{r}
ggplot(aes(x= citric.acid), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous() +
  xlab("Citric Acid (g / dm^3)") + ylab("Count")

```

##### I chose a bar chart so we can easily the citric acid content of wines in the dataset.

#### Box Plot with Category Summaries
```{r cacid, warning=FALSE}

ggplot(aes(factor(quality), 
            citric.acid), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Citric Acid (g / dm^3)") + ylab("Count")

```

#### This seems in line with our analysis of fixed acidity. Higher ratings tend to have higher citric acid content.

### Residual Sugar
#### Bar Chart
```{r}
ggplot(aes(x= residual.sugar), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous() +
  xlab("Residual Sugar (g / dm^3)") + ylab("Count")

```

##### I chose a bar chart so we can easily the residual sugar content of wines in the dataset.

#### Box Plot with Category Summaries
```{r rsugar, warning=FALSE}

ggplot(aes(factor(quality), 
            residual.sugar), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Residual Sugar (g / dm^3)") + ylab("Count")

```

#### It seems as if risidual sugar has minimal impact on quality ratings.

### pH
#### Bar Chart
```{r}
ggplot(aes(x= pH), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous() +
  xlab("pH") + ylab("Count")

```

##### I chose a bar chart so we can easily the pH of wines in the dataset.

#### Box Plot with Category Summaries
```{r pH, warning=FALSE}

ggplot(aes(factor(quality), 
            pH), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("pH") + ylab("Count")

```

#### pH is a measure of acidity. The lower the pH, the higher the acidity. This is also in line with our findings that higher rated wines tend to have a higher acidity (or lower pH as we see here).

### Total Sulfur Dioxide Content
#### Bar Chart

```{r}
ggplot(aes(x= total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous()+
  xlab("Sulfur Dioxide (mg / dm^3)") + ylab("Count")

```

##### I chose a bar chart so we can easily the sulfur dioxide content of wines in the dataset.
#### Box Plot with Category Summaries
```{r suldioxhist, warning=FALSE}

ggplot(aes(factor(quality), 
            total.sulfur.dioxide), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Total Sulfur Dioxide (mg / dm^3)") + ylab("Count")

```

#### Interesting, let's dig in and take a look at free sulfur dioxide

### Free Sulfur Dioxide and Wine Quality

#### Bar Chart
```{r}
ggplot(aes(x= free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous()+
  xlab("Free Sulfur Dioxide (mg / dm^3)") + ylab("Count")

```

##### I chose a bar chart so we can easily the free sulfur dioxide content of wines in the dataset.
#### Box Plot with Category Summaries

```{r freesuldioxhist, warning=FALSE}

ggplot(aes(factor(quality), 
            free.sulfur.dioxide), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  xlab("Free Sulfur Dioxide (mg / dm^3)") + ylab("Count")

```

#### This is very similar to the previous plot looking at total sulfur dioxide. THis would be a good rabbit hole to follow in a further analysis.




### Total Sulfur Dioxide and Wine Quality
#### Sulfur Dioxide is used to preserve wine. There are many different views on its use in wine. I am interested in whether the amount of sulfur dioxide impacts quality ratings.

```{r sulfurdioxQuality, warning=FALSE}
ggplot(aes(x = quality, y = total.sulfur.dioxide), data = wine) +
  geom_jitter(alpha = 1/20) +
  coord_trans(y = 'sqrt')+
  xlab("Quality Rating") + ylab("Total Sulfur Dioxide (mg / dm^3)")

summary(wine$total.sulfur.dioxide)

```

#### This scatterplot shows that there is a tendency of wines with 3 or 4 rating to contain more sulfur dioxide than those with a higher rating.

### Alcohol and Wine Quality
#### I am interested in whether the amount of alcohol in wine impacts quality ratings.

```{r alcoholQuality, warning=FALSE}
ggplot(aes(x = quality, y = alcohol), data = wine) +
  geom_jitter(alpha = 1/15) +
  coord_trans(y = 'sqrt')+
  xlab("Quality Rating") + ylab("Alcohol (% by volume)")

summary(wine$alcohol)

```

#### This scatterplot shows that there is a tendency of wines with a higher rating to contain more alcohol.

### Residual Sugars and Wine Quality
#### I am interested in whether the amount of residual sugar in wine impacts quality ratings.

```{r sugarqual, warning=FALSE}
ggplot(aes(x = quality, y = residual.sugar), data = wine) +
  geom_jitter(alpha = 1/15) +
  coord_trans(y = 'sqrt')+
  xlab("Quality Rating") + ylab("Residual Sugar (g / dm^3)")

```

### pH and Wine Quality
#### I am interested in whether the pH of wine impacts quality ratings.

```{r pHqual, warning=FALSE}
ggplot(aes(x = quality, y = pH), data = wine) +
  geom_jitter(alpha = 1/15) +
  coord_trans(y = 'sqrt')+
  xlab("Quality Rating") + ylab("pH")

```

### Alcohol & Sugar's Impact on Quality

```{r alcoholSugarqual, warning=FALSE}
ggplot(aes(x = alcohol, 
           y = residual.sugar  , color = factor(quality)), 
       data = wine) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE, size=1)  +
  scale_color_brewer(palette = 3, type='seq',
                   guide=guide_legend(title='Quality'))+
  xlab("Alcohol (% by volume)") + ylab("residual sugar (g / dm^3)")
```

#### This multivariate plot shows the impact of alcohol content and residual sugar's impact on quality. We can see that a wine with lower residual sugar content has a higher rating, which increases as alcohol content increases. There is a sharper rise in quality as lower rated wines increase in alcohol content.

### Chlorides & Sulphates' Impact on Quality

```{r chlorSulphqual, warning=FALSE}
ggplot(aes(x = sulphates, 
           y = chlorides  , color = factor(quality)), 
       data = wine) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(palette = 3, type='seq',
                   guide=guide_legend(title='Quality'))+
  xlab("sulphates (potassium sulphate - g / dm3)") + ylab("chlorides (sodium chloride - g / dm^3)")
```

#### High rated wines tend to have a lower number of sulphates and chlorides. As sulphates decrease the wine's quality increases.

### At what point do sulphates impact quality given alcohol content
#### Let's determine our bucket size by looking at the sulphates data.

```{r}
summary(wine$sulphates)

```

#### WE'll set our buckets based on the quartiles.

```{r sulbox, warning=FALSE}
wine$sulphates.bucket = cut(wine$sulphates,
                            c(.30, .55, .62, .73, 1))

ggplot(aes(x = factor(quality),
           y = sulphates ), 
           data = wine) + 
   geom_boxplot( aes(fill= sulphates.bucket))  +
  scale_fill_brewer(type='seq',
                   guide=guide_legend())+
  xlab("Quality") + ylab("sulphates (potassium sulphate - g / dm3)")

```

## Final Plots and Summary

##### The following charts represent the most relevant findings of this analysis.

### Alcohol Content
```{r fnlalcohol, warning=FALSE}

ggplot(aes(factor(quality), 
            alcohol), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  xlab("Quality") + ylab("alcohol (% by volume)")

```

###### The higher the alcohol content the higher the rating.

### Fixed Acidity
```{r fnlfacid, warning=FALSE}

ggplot(aes(factor(quality), 
            fixed.acidity), 
        data = wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  xlab("Quality") + ylab("fixed acidity (tartaric acid - g / dm^3)")

```

###### The higher the rating, the higher the fixed acidity of the wine.

### Chlorides & Sulphates' Impact on Quality

```{r fnlchlorSulphqual, warning=FALSE}
ggplot(aes(x = sulphates, 
           y = chlorides  , color = factor(quality)), 
       data = wine) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(palette = 3, type='seq',
                   guide=guide_legend(title='Quality'))+
  xlab("sulphates (potassium sulphate - g / dm3)") + ylab("chlorides (sodium chloride - g / dm^3")
```

#### High rated wines tend to have a lower number of sulphates and chlorides. As sulphates decrease the wine's quality increases.

## Reflection

#### The analysis above attempts to determine if there is a coorelation between alcohol content and quality and total sulfur dioxide content and quality.

#### I plotted the alcohol variable against the quality variable to determine that a higher alcohol content does coorelate with a higher quality rating.

#### I plotted the total Sulfur Dioxide variable against quality and found that lower sulfur dioxide content coorelates with a higher quality rating.

#### For further analysis, I would question whether or not acidity affects wine quality ratings.